---
title: "Exercise 2_1: Vector Data Refresher"
subtitle: 'Advanced Geospatial Data Processing for Social Scientists'
author: Dennis Abel & Stefan Jünger
format: 
  html:
    embed-resources: true
    code-copy: true
execute:
  echo: true
editor: visual
callout-icon: false
callout-appearance: minimal
editor_options: 
  chunk_output_type: console
---

In these three exercises, you’ll refresh your knowledge about geospatial vector data using `R`. We’ll start with some basics on importing and handling vector data, and then explore how to link datasets together — even when the data doesn’t always match perfectly.

Along the way, you’ll have the chance to explore, experiment, and solve problems that come up, like working with different coordinate reference systems (CRS).

Let’s get started and have some fun with spatial data! 🌍✨

### Exercises

::: callout-note
## 🏋️ Exercise 1

Import a polygon vector shapefile from the `./data` folder. After loading, explore it:

- How many features are there? What attributes are included?
- What is the coordinate reference system (CRS)?
- Have a look at the `GF` column — it contains different values. What could they mean, and how many unique geometries do you actually have?
- Feel free to plot the data for a first visual impression!
:::

::: {.callout-caution collapse="true"}
## 💡 Tip

Shapefiles include a file with the extension `.shp`. You can use the `sf` package and the `sf::st_read()` function to import the data. Hint: If you want a cleaner output, you can set `quiet = TRUE`.

The `GF` column encodes the geometry type (e.g., land vs. water boundaries). Try `table(data$GF)` to see the distribution — you may want to filter for one type!
:::

::: callout-note
## 🏋️ Exercise 2

Import the file `synthetic_coordinates.csv` from the `./data` folder. It contains point coordinates, but it’s not yet an `sf` object! After loading, convert it into a proper spatial object by assigning coordinates and a CRS. (Feel free to plot the points afterward if you like!)
:::

::: {.callout-caution collapse="true"}
## 💡 Tip

First, read the table using `read.csv()` or `readr::read_csv()`. Then use `sf::st_as_sf()` to create a spatial object, specifying the coordinate columns and CRS. Hint: check the `crs` column in the CSV — it tells you which EPSG code to use!
:::

::: callout-note
## 🏋️ Exercise 3

Let’s connect the two datasets you imported! Join the points from `synthetic_coordinates.csv` with the polygons you loaded earlier, adding the `AGS` column from the polygon dataset to the points.
:::

::: {.callout-caution collapse="true"}
## 💡 Tip

Beware: The coordinate reference systems (CRS) of the two datasets are different! Make sure they match before trying to perform the spatial join — otherwise, you might get an error (and that’s totally fine — part of the learning!).

- Check the CRS of both datasets with `sf::st_crs()`.
- Use `sf::st_transform()` to bring one dataset into the CRS of the other.
- For the join, you can use `sf::st_join()`.

You could also transform the polygons instead of the points — it depends on what fits better. Always double-check that geometries line up properly after transformation (e.g., by plotting)!
:::

### Solutions

::: {.callout-tip collapse="true"}
## ✅ Solution 1

```{r}
#| eval: false
library(sf)

# Load the German districts shapefile
polygon_data <- sf::st_read("./data/VG250_KRS.shp", quiet = TRUE)

# Basic inspection — how big is the dataset?
print(polygon_data) # overview: columns, geometry type, CRS
nrow(polygon_data) # number of features (rows)
sf::st_crs(polygon_data) # coordinate reference system

# The GF column contains geometry type codes
# GF = 4 means land boundaries (what we usually want)
table(polygon_data$GF) # how many of each type?

# Visualize the geometry (without attributes)
plot(sf::st_geometry(polygon_data))
```

```{r}
#| echo: false
library(sf)

polygon_data <- sf::st_read("../../data/VG250_KRS.shp", quiet = TRUE)

print(polygon_data)
nrow(polygon_data)
sf::st_crs(polygon_data)

table(polygon_data$GF)

plot(sf::st_geometry(polygon_data))
```
:::

::: {.callout-tip collapse="true"}
## ✅ Solution 2

```{r}
#| eval: false
library(sf)
library(readr)

# Load as plain data frame — no geometry yet!
coords_df <- read_csv("./data/synthetic_geocoordinates.csv")
head(coords_df) # inspect: which columns hold coordinates?

# Convert to sf: specify coordinate columns and CRS
# The x/y columns are in EPSG:3035 (ETRS89-LAEA, meters)
points_sf <-
  sf::st_as_sf(
    coords_df,
    coords = c("x", "y"), # coordinate columns
    crs = 3035 # ETRS89-LAEA (Europe equal-area)
  )

# Now it's a proper spatial object
plot(points_sf)
```

```{r}
#| echo: false
library(sf)
library(readr)

coords_df <- readr::read_csv("../../data/synthetic_geocoordinates.csv")

points_sf <-
  sf::st_as_sf(
    coords_df,
    coords = c("x", "y"),
    crs = 3035
  )

plot(points_sf)
```
:::

::: {.callout-tip collapse="true"}
## ✅ Solution 3

```{r}
# Step 1: Check CRS of both datasets — they must match for a join!
sf::st_crs(points_sf)$epsg # points: EPSG 3035 (ETRS89-LAEA)
sf::st_crs(polygon_data)$epsg # polygons: different CRS

# Step 2: Reproject points to match the polygon CRS
# (We transform the points because they're lighter than polygons)
points_transformed <-
  sf::st_transform(points_sf, crs = sf::st_crs(polygon_data))

# Step 3: Spatial join — adds AGS (district code) to each point
# based on which polygon it falls into
points_with_AGS <- sf::st_join(points_transformed, polygon_data[, "AGS"])

# Check: each point should now have an AGS value
head(points_with_AGS)
```
:::
