---
title: "Exercise 3_2: Extracting and Analyzing Raster Information"
subtitle: 'Advanced Geospatial Data Processing for Social Scientists'
author: Dennis Abel & Stefan Jünger
format: 
  html:
    embed-resources: true
    code-copy: true
execute:
  echo: true
knitr:
  opts_knit:
    root.dir: ../../
editor: visual
callout-icon: false
callout-appearance: minimal
editor_options: 
  chunk_output_type: console
---

In this set of exercises, you'll focus on linking raster data, aggregating them to a coarser resolution, creating heatmaps using spatial smoothing, and linking processed raster data with your synthetic geocoordinates. You will:

- Aggregate raster data to a larger grid (e.g., 5 km resolution).
- Use a focal operation to create a heatmap and visualize spatial patterns.
- Link the aggregated raster back to your geocoordinates for further analysis.

You'll work with Census 2022 rasters from `./data/z22/`. Not sure what a file contains? File names match Census variable names — check the [Census 2022 database](https://ergebnisse.zensus2022.de/datenbank/online/themes) for documentation, or use `z22::z22_features()` if you have the [`z22` package](https://cran.r-project.org/package=z22) installed.

### Exercises

::: callout-note
## 🏋️ Exercise 1

Your first task is to import a dataset of synthetic geographic coordinates (`synthetic_survey_geocoordinates.rds` in `./data/survey/`) and a raster `.tif` dataset from the `./data/z22/` folder.

After loading both, compare their coordinate reference systems. Do they match? What would you need to do before extracting raster values at the point locations?
:::

::: {.callout-caution collapse="true"}
## 💡 Tip

- Use `readRDS()` to load the `.rds` file and `terra::rast()` for the raster.
- Check the CRS of the points with `sf::st_crs()` and of the raster with `terra::crs(..., describe = TRUE)`.
- After loading, inspect the structure with `str()` and visualize with `plot()`.
:::

::: callout-note
## 🏋️ Exercise 2

In this exercise, you’ll extract raster values at the locations of the synthetic coordinates you loaded previously. But here’s the twist: Think about the surrounding area!

Use buffers or a spatial filter (like `terra::focal()`) to capture information around each point, not just at the point itself. This will help you get a broader view of the raster values in the neighborhood of your coordinates.

Important: You may notice that the CRS of the points and the CRS of the raster don’t align. You will need to adjust the CRS of your points to match the raster's CRS to correctly extract values.
:::

::: {.callout-caution collapse="true"}
## 💡 Tip

- You can use `terra::extract()` to pull raster values at specific coordinates.
- To consider the surrounding area, use `terra::focal()` to apply a spatial filter or create buffers with `sf::st_buffer()` around your points.
- Be creative: What size buffer makes sense for your data? Should you use a large or small neighborhood around the points? How do the values change within this window?
:::

::: callout-note
## 🏋️ Exercise 3

In this exercise, you’ll explore how spatial resolution affects your results.

- Aggregate the raster to **two different resolutions** (e.g., a 5 km and a 10 km grid).
- Apply a focal smoothing operation to both aggregated rasters to create heatmaps.
- Extract values at your synthetic coordinates from both versions.
- Compare the extracted values: How much do they differ? Plot both heatmaps side by side. At which resolution do you start losing important spatial patterns?
:::

::: {.callout-caution collapse="true"}
## 💡 Tip

- To aggregate raster data, use `terra::aggregate()` with different `fact` values (e.g., `fact = 5` for 5 km, `fact = 10` for 10 km).
- Use `terra::focal()` to apply a moving window average for the heatmap effect.
- Use `par(mfrow = c(1, 2))` to plot both heatmaps side by side.
- After aggregation and smoothing, link both rasters back to your coordinates using `terra::extract()` and compare.
:::

### Solutions

::: {.callout-tip collapse="true"}
## ✅ Solution 1

```{r}
library(terra)

# Load the synthetic survey coordinates (an sf point dataset)
coordinates <- readRDS("./data/survey/synthetic_survey_geocoordinates.rds")

# Inspect: how many points? which columns? which geometry type?
str(coordinates)

# Load a raster dataset — pick any .tif from ./data/z22/
rent_avg <- terra::rast("./data/z22/rent_avg.tif")

# Visualize the raster to get a first impression
plot(rent_avg)

# Key step: compare the CRS of both datasets
# Points use EPSG from the survey, raster uses the Census grid CRS
sf::st_crs(coordinates)$epsg # CRS of the point data
terra::crs(rent_avg, describe = TRUE)$code # CRS of the raster

# They don't match! Before extracting raster values at point locations,
# we'll need to reproject one to match the other using sf::st_transform().
```
:::

::: {.callout-tip collapse="true"}
## ✅ Solution 2

```{r}
library(terra)

# Step 1: Fix the CRS mismatch we found in Exercise 1
# We reproject the points to match the raster's CRS
crs_points <- sf::st_crs(coordinates)$epsg
crs_raster <- terra::crs(rent_avg, describe = TRUE)$code

if (crs_points != crs_raster) {
  coordinates <- sf::st_transform(coordinates, crs = as.numeric(crs_raster))
}

# Step 2: Create buffers around each point
# 5000 = 5 km in projected CRS (units are meters)
buffer <- sf::st_buffer(coordinates, 5000)

# Quick check: do buffers look reasonable?
plot(sf::st_geometry(buffer))
plot(rent_avg)

# Step 3a: Extract raster values directly at point locations (1:1)
rent_buffer_values <- terra::extract(rent_avg, coordinates)

# Step 3b: Alternative — smooth the raster first with a focal window,
# then extract. This captures the neighborhood context.
rent_focal <-
  terra::focal(rent_avg, w = matrix(1, 5, 5), fun = mean, na.rm = TRUE)

rent_focal_values <-
  terra::extract(rent_focal, coordinates)

# Compare: point extraction vs. focal-smoothed extraction
# The focal values should be less extreme (smoothed towards neighbors)
head(rent_buffer_values)
head(rent_focal_values)
```
:::

::: {.callout-tip collapse="true"}
## ✅ Solution 3

```{r}
# Step 1: Aggregate the original raster to two resolutions
# fact = 5 combines 5x5 cells → ~5 km; fact = 10 → ~10 km
rent_5km <- terra::aggregate(rent_avg, fact = 5, na.rm = TRUE)
rent_10km <- terra::aggregate(rent_avg, fact = 10, na.rm = TRUE)

# Step 2: Smooth both with a focal window to create heatmaps
# Same window size (5x5), but applied to different base resolutions
rent_heatmap_5km <-
  terra::focal(rent_5km, w = matrix(1, 5, 5), fun = mean, na.rm = TRUE)
rent_heatmap_10km <-
  terra::focal(rent_10km, w = matrix(1, 5, 5), fun = mean, na.rm = TRUE)

# Step 3: Visual comparison — plot side by side
# Notice how the 10 km version loses local patterns (e.g., city centers)
par(mfrow = c(1, 2))
plot(rent_heatmap_5km, main = "Heatmap (5 km)")
plot(rent_heatmap_10km, main = "Heatmap (10 km)")
par(mfrow = c(1, 1))

# Step 4: Extract values at our survey coordinates from both
values_5km <- terra::extract(rent_heatmap_5km, coordinates)
values_10km <- terra::extract(rent_heatmap_10km, coordinates)

# Inspect: the 10 km values should be more uniform (less variation)
head(values_5km)
head(values_10km)

# Quantify the difference: mean absolute difference per point
# Higher values = more information lost at coarser resolution
mean(abs(values_5km[[2]] - values_10km[[2]]), na.rm = TRUE)
```
:::
