---
title: "Exercise 7_1: Data Integration"
subtitle: 'Advanced Geospatial Data Processing for Social Scientists'
author: Dennis Abel & Stefan Jünger
format:
  html:
    embed-resources: true
    code-copy: true
execute:
  echo: true
knitr:
  opts_knit:
    root.dir: ../../
editor: visual
callout-icon: false
callout-appearance: minimal
editor_options:
  chunk_output_type: console
---

::: callout-note
## 🏡 Take-home exercise

This exercise is designed for **after the course** — work on it at your own pace. It brings together everything you've learned across all sessions.
:::

In this session, we walked through a full research workflow — from accessing Earth observation data to linking it with survey data. Now it's your turn! Choose one of two paths:

- **Path A** is open-ended: formulate your own spatial research question and tell a "spatial story" using the course data.
- **Path B** provides a structured step-by-step exercise if you prefer clear guidance.

Both paths practice the core skill of this session: **combining multi-layer geospatial data with tabular data through spatial aggregation and joins.**

## Path A: Spatial Story (open-ended) {.unnumbered}

::: callout-note
## 🏋️ Your task

**Tell a spatial story using at least two data sources from this course.**

Formulate a research question and answer it with data. Your analysis must:

1.  Use at least one **multi-layer dataset** (a `.nc` Census cube, a raster stack, or a time series)
2.  Include a **linkage** between data sources (spatial join, raster extraction, or tabular join)
3.  Produce at least one **visualization** of the result

Prepare 2–3 sentences interpreting what you found — and whether you trust the result.
:::

::: {.callout-caution collapse="true"}
## 💡 Inspiration

Here are some questions you could explore (or invent your own!):

- **Age structure × nighttime lights**: Does the age distribution (from `age_short.nc`) differ between bright and dark areas? Extract nighttime lights and age groups per district, compare.
- **Housing × population change**: Is there a relationship between vacancy rates (`vacancies.tif`) and population trends from the WorldPop time series (2017–2020)?
- **Citizenship × survey context**: Extract citizenship composition (`citizenship.nc`) around the synthetic survey coordinates. Does the neighborhood composition differ by respondents' citizenship status?
- **Heating × rent**: Do districts with more renewable heating (`building_heat_src.nc`) have higher or lower rents (`rent_avg.tif`)?
- **Your own data**: If you brought your own dataset, this is the time to try integrating it with the Census or WorldPop data!

Remember: Layer names in `.nc` files can be cryptic — use `z22::z22_categories()` or the [Census 2022 database](https://ergebnisse.zensus2022.de/datenbank/online/themes) to decode them.
:::

## Path B: Neighborhood profiles from Census cubes (structured) {.unnumbered}

In this exercise, you'll build **neighborhood context profiles** from Census data cubes and link them to the synthetic survey respondents. This mirrors the session workflow — but at a finer spatial scale and with multi-layer data.

### Exercises

::: callout-note
## 🏋️ Exercise 1 — Explore a Census data cube

Load a `.nc` file from `./data/z22/` — for example, `age_short.nc`. Inspect it:

- How many layers does it have?
- What do the layer names mean? Decode them using `z22::z22_categories("age_short")` or the [Census 2022 database](https://ergebnisse.zensus2022.de/datenbank/online/themes).
- Plot the individual layers side by side. What spatial patterns do you see?
:::

::: {.callout-caution collapse="true"}
## 💡 Tip

`terra::rast()` handles `.nc` files just like `.tif` — but returns multiple layers. Use `terra::nlyr()` to count them, `names()` to see their names, and `terra::plot()` to visualize all layers at once.
:::

::: callout-note
## 🏋️ Exercise 2 — Extract neighborhood context for survey respondents

Load the synthetic survey coordinates (`synthetic_survey_geocoordinates.rds`). Extract the values of **all layers** of your Census cube at the respondent locations.

Do this twice:

1.  **Point extraction** — direct 1:1 extraction at each coordinate
2.  **Buffer extraction** — using a 5 km buffer around each point, averaging across cells

Compare the results: How much do the two approaches differ? Why might you prefer one over the other in practice?
:::

::: {.callout-caution collapse="true"}
## 💡 Tip

- Watch out for CRS mismatches! Check `sf::st_crs()` and `terra::crs()` before extracting.
- For the buffer approach, use `sf::st_buffer(points, 5000)` (assuming meters) and pass it to `terra::extract(..., fun = mean, na.rm = TRUE)`.
- `terra::extract()` works on multi-layer rasters — it returns one column per layer automatically.
:::

::: callout-note
## 🏋️ Exercise 3 — Link and compare groups

The synthetic survey coordinates contain an attribute on citizenship status. Use it to compare the **neighborhood profiles** between the two groups:

- Compute the mean of each Census cube layer per group (citizenship status × age bracket)
- Visualize the difference — for example, as a grouped barplot of age shares by citizenship status
- Discuss: Does the spatial context differ between groups? And: since the data are synthetic, can we interpret this as a "real" finding — or is it an artifact?
:::

::: {.callout-caution collapse="true"}
## 💡 Tip

- After `terra::extract()`, bind the results to the citizenship column — not the full points dataset! The points have other columns with different types that will cause problems during pivoting. Use e.g. `dplyr::tibble(citizenship = points$citizenship) |> dplyr::bind_cols(extracted_values)`.
- Use `tidyr::pivot_longer()` to reshape the wide extraction results into a long format suitable for `ggplot2`.
- For the barplot, `ggplot2::geom_col(position = "dodge")` with `facet_wrap()` by age group works well.
- The critical reflection is part of the exercise — synthetic data are designed to have spatial structure, so patterns you find are intentional but not empirically "real".
:::

### Solutions

::: {.callout-tip collapse="true"}
## ✅ Solution 1

```{r}
library(terra)

# Load the Census data cube
age_cube <- terra::rast("./data/z22/age_short.nc")

# How many layers?
terra::nlyr(age_cube)

# What are they called?
names(age_cube)

# Decode the layer names — what does each layer represent?
# Option 1: z22 package (if installed)
# z22::z22_categories("age_short")
# Option 2: check https://ergebnisse.zensus2022.de/datenbank/online/themes

# Inspect the metadata
print(age_cube)

# Plot all layers — each shows a different age group's share
terra::plot(age_cube)
```
:::

::: {.callout-tip collapse="true"}
## ✅ Solution 2

```{r}
library(sf)

# Load survey coordinates
points <- readRDS("./data/survey/synthetic_survey_geocoordinates.rds")

# Check CRS of both datasets
sf::st_crs(points)$epsg
terra::crs(age_cube, describe = TRUE)$code

# Reproject points to match the raster CRS
points <- sf::st_transform(points, terra::crs(age_cube))

# --- Point extraction (1:1) ---
extracted_point <- terra::extract(age_cube, points, ID = FALSE)
head(extracted_point)

# --- Buffer extraction (5 km neighborhood) ---
buffers <- sf::st_buffer(points, 5000)
extracted_buffer <- terra::extract(
  age_cube, buffers, fun = mean, na.rm = TRUE, ID = FALSE
)
head(extracted_buffer)

# --- Compare the two approaches ---
# The buffer values should be smoother (less extreme)
# because they average over multiple grid cells
summary(extracted_point)
summary(extracted_buffer)
```
:::

::: {.callout-tip collapse="true"}
## ✅ Solution 3

```{r}
library(dplyr)
library(tidyr)
library(ggplot2)

# Combine extraction results with the citizenship attribute
points_enriched <-
  dplyr::tibble(citizenship = points$citizenship) |>  # keep only grouping variable
  dplyr::bind_cols(extracted_buffer) |>               # add extracted values
  tidyr::pivot_longer(                                # reshape to long format
    cols = -citizenship,                              # pivot only the age layers
    names_to = "age_group",
    values_to = "share"
  )

# Compute group means
group_profiles <- points_enriched |>
  dplyr::group_by(citizenship, age_group) |>
  dplyr::summarize(mean_share = mean(share, na.rm = TRUE), .groups = "drop")

# Visualize
ggplot(group_profiles, aes(x = age_group, y = mean_share, fill = factor(citizenship))) +
  geom_col(position = "dodge") +
  labs(
    title = "Neighborhood age profiles by citizenship status",
    x = "Age group (Census cube layer)",
    y = "Mean share in 5 km buffer",
    fill = "Citizenship\nstatus"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

The synthetic data were designed to have spatial structure, so any pattern we find is "built in" by the data generation process — not evidence for a real-world association. In a real analysis, this is exactly where you'd discuss ecological fallacy, selection effects, and whether the spatial scale of your buffer matches your theoretical mechanism.
:::
